---
title: "Control vs VPA-Treated Dorsal Forebrain Organoids: scRNA-seq Analysis"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: now
date-format: "YYYY-MM-DD HH:mm"
format:
  html:
    page-layout: full
    embed-resources: true
    toc: true
    code-fold: true
    code-summary: "Show code"
    include-in-header:
      text: |
        <style>
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
        </style>
---

## Introduction

Valproic acid (VPA) is a widely prescribed antiepileptic drug that also acts as a
**histone deacetylase (HDAC) inhibitor**. Prenatal exposure to VPA is an established
teratogen and has been linked to increased risk of autism spectrum disorder and other
neurodevelopmental conditions.

This analysis examines single-cell RNA-seq data from
[Yentür et al. (2025, bioRxiv)](https://doi.org/10.1101/2025.01.09.632137),
comparing **control** vs **VPA-treated** human dorsal forebrain organoids.
The dataset was obtained from
[CELLxGENE](https://cellxgene.cziscience.com/) and contains 26,499 cells across
5 annotated cell types.

**Experimental design**: 2 donors, each contributing organoids to both control and
VPA conditions (a paired design). Because we have only 2 biological replicates per
condition, all differential expression results should be considered **exploratory**.

**What this analysis covers**:

1. Visualization of cell populations (UMAP)
2. Cell type composition changes under VPA
3. Pseudobulk differential expression per cell type (limma-voom, paired design)
4. Gene set enrichment analysis (GSEA with Hallmark and GO gene sets)
5. Expression of known VPA-response and neural development genes


## Setup & Data Loading

```{r}
#| label: setup
#| message: false
#| warning: false
library(SingleCellExperiment)
library(SummarizedExperiment)
library(edgeR)
library(limma)
library(ggplot2)
library(plotly)
library(patchwork)
library(DT)
library(dplyr)
library(tidyr)
library(ggrepel)
library(fgsea)
library(msigdbr)
library(pheatmap)
library(scales)
library(glue)
library(here)
library(Matrix)
library(viridis)

base_size <- 14
theme_set(theme_bw(base_size = base_size))
```

```{r}
#| label: load-data
sce <- readRDS(here::here("analysis", "vpa_organoids_sce.rds"))

# Map Ensembl IDs to gene symbols where available
# feature_name contains gene symbols for annotated genes, ENSG IDs otherwise
gene_symbols <- as.character(rowData(sce)$feature_name)
is_symbol <- !grepl("^ENSG", gene_symbols)

# Make symbols unique (some may be duplicated)
symbol_map <- gene_symbols
symbol_map[is_symbol] <- make.unique(symbol_map[is_symbol])
names(symbol_map) <- rownames(sce)

# Create a lookup from ENSG to symbol
ensg_to_symbol <- setNames(gene_symbols, rownames(sce))

# Rename rows to gene symbols where available
rownames(sce) <- symbol_map
```

The dataset contains **`r format(ncol(sce), big.mark = ",")`
cells** and **`r format(nrow(sce), big.mark = ",")`
genes** (`r sum(is_symbol)` with gene symbols, `r sum(!is_symbol)` Ensembl IDs only).

```{r}
#| label: color-palettes
cell_type_colors <- c(
  "progenitor cell" = "#1b9e77",
  "neuronal-restricted precursor" = "#d95f02",
  "cerebral cortex neuron" = "#7570b3",
  "Cajal-Retzius cell" = "#e7298a",
  "unknown" = "#999999"
)

condition_colors <- c(
  "CTRL" = "#4393c3",
  "VPA" = "#d6604d"
)

donor_colors <- c(
  "#66c2a5",
  "#fc8d62"
)
names(donor_colors) <- sort(unique(sce$donor_id))
```

```{r}
#| label: gene-link-helper
#' Create a clickable GeneCards link for a gene symbol
gene_link <- function(symbol) {
  paste0(
    '<a href="https://www.genecards.org/cgi-bin/carddisp.pl?gene=',
    symbol,
    '" target="_blank">',
    symbol,
    '</a>'
  )
}
```


## Experimental Design Overview

The experiment uses a **paired design**: each donor contributes cells to both
conditions. This lets us account for donor-to-donor variability when testing
for VPA effects.

```{r}
#| label: design-tables
knitr::kable(
  as.data.frame.matrix(table(sce$condition, sce$donor_id)),
  caption = "Cells per condition × donor"
)

knitr::kable(
  as.data.frame.matrix(table(sce$condition, sce$cell_type)),
  caption = "Cells per condition × cell type"
)
```

```{r}
#| label: design-table-dt
design_df <- as.data.frame(
  table(
    Condition = sce$condition,
    Donor = sce$donor_id,
    CellType = sce$cell_type
  )
) |>
  dplyr::filter(Freq > 0) |>
  tidyr::pivot_wider(
    names_from = Condition,
    values_from = Freq,
    values_fill = 0
  )

DT::datatable(
  design_df,
  caption = "Cell counts by donor, condition, and cell type",
  options = list(pageLength = 20)
)
```


## UMAP Visualizations

UMAP (Uniform Manifold Approximation and Projection) is a dimensionality reduction
technique that projects high-dimensional gene expression data into 2D for
visualization. Cells that are close together on the UMAP have similar expression
profiles. The UMAP coordinates here were computed by the original study authors.

```{r}
#| label: umap-data
umap_df <- data.frame(
  UMAP1 = reducedDim(sce, "X_umap")[, 1],
  UMAP2 = reducedDim(sce, "X_umap")[, 2],
  cell_type = sce$cell_type,
  condition = sce$condition,
  donor_id = sce$donor_id,
  cell_phase = sce$cell_phase
)
```

### UMAP by Cell Type

```{r}
#| label: umap-cell-type
#| fig-width: 9
#| fig-height: 7
ggplot(umap_df, aes(UMAP1, UMAP2, color = cell_type)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_color_manual(values = cell_type_colors) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title = "UMAP colored by cell type", color = "Cell type")
```

### UMAP by Condition

```{r}
#| label: umap-condition
#| fig-width: 9
#| fig-height: 7
ggplot(umap_df, aes(UMAP1, UMAP2, color = condition)) +
  geom_point(size = 0.3, alpha = 0.4) +
  scale_color_manual(values = condition_colors) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title = "UMAP colored by condition", color = "Condition")
```

### UMAP Split by Condition

```{r}
#| label: umap-split-condition
#| fig-width: 14
#| fig-height: 6
ggplot(umap_df, aes(UMAP1, UMAP2, color = cell_type)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_color_manual(values = cell_type_colors) +
  facet_wrap(~condition) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title = "UMAP split by condition", color = "Cell type")
```

### UMAP by Donor

```{r}
#| label: umap-donor
#| fig-width: 9
#| fig-height: 7
ggplot(umap_df, aes(UMAP1, UMAP2, color = donor_id)) +
  geom_point(size = 0.3, alpha = 0.4) +
  scale_color_manual(values = donor_colors) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title = "UMAP colored by donor", color = "Donor")
```

### UMAP by Cell Cycle Phase

```{r}
#| label: umap-cell-phase
#| fig-width: 9
#| fig-height: 7
ggplot(umap_df, aes(UMAP1, UMAP2, color = cell_phase)) +
  geom_point(size = 0.3, alpha = 0.4) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title = "UMAP colored by cell cycle phase", color = "Cell phase")
```

### Interactive UMAP (Plotly)

:::{.column-screen}
```{r}
#| label: umap-plotly
#| fig-width: 10
#| fig-height: 7
# Subsample for plotly performance
set.seed(42)
idx <- sample(nrow(umap_df), min(10000, nrow(umap_df)))
umap_sub <- umap_df[idx, ]

p <- plot_ly(
  umap_sub,
  x = ~UMAP1,
  y = ~UMAP2,
  color = ~cell_type,
  colors = cell_type_colors,
  text = ~ paste(
    "Cell type:",
    cell_type,
    "<br>Condition:",
    condition,
    "<br>Donor:",
    donor_id
  ),
  type = "scatter",
  mode = "markers",
  marker = list(size = 3, opacity = 0.6),
  width = 900,
  height = 650
) |>
  layout(
    title = "Interactive UMAP (10,000 cell subsample)",
    xaxis = list(title = "UMAP1"),
    yaxis = list(title = "UMAP2")
  )
p

```
:::


## Cell Type Composition Analysis

A key question in case-control scRNA-seq studies is whether the **proportions** of
cell types shift between conditions. Changes in composition can reflect altered
differentiation, proliferation, or survival.

```{r}
#| label: composition-data
comp_df <- as.data.frame(table(
  condition = sce$condition,
  cell_type = sce$cell_type,
  donor_id  = sce$donor_id
))

# Compute proportions within each condition × donor
comp_df <- comp_df |>
  group_by(condition, donor_id) |>
  mutate(
    total = sum(Freq),
    proportion = Freq / total
  ) |>
  ungroup()
```

### Stacked Bar Chart

```{r}
#| label: composition-stacked
#| fig-width: 10
#| fig-height: 6
# Per condition (aggregated across donors)
comp_cond <- comp_df |>
  group_by(condition, cell_type) |>
  summarize(Freq = sum(Freq), .groups = "drop") |>
  group_by(condition) |>
  mutate(proportion = Freq / sum(Freq)) |>
  ungroup()

ggplot(comp_cond, aes(condition, proportion, fill = cell_type)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = cell_type_colors) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cell type composition by condition",
    x = "Condition", y = "Proportion", fill = "Cell type"
  )
```

### Composition by Donor

```{r}
#| label: composition-by-donor
#| fig-width: 12
#| fig-height: 6
comp_df$sample <- paste(comp_df$condition, comp_df$donor_id, sep = "\n")

ggplot(comp_df, aes(sample, proportion, fill = cell_type)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = cell_type_colors) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cell type composition by condition and donor",
    x = "Sample", y = "Proportion", fill = "Cell type"
  )
```

### Composition Table

```{r}
#| label: composition-table
comp_table <- comp_cond |>
  mutate(
    label = paste0(Freq, " (", round(proportion * 100, 1), "%)")
  ) |>
  dplyr::select(condition, cell_type, label) |>
  tidyr::pivot_wider(names_from = condition, values_from = label)

DT::datatable(comp_table, caption = "Cell counts and proportions per condition",
              escape = FALSE, options = list(dom = "t"))
```


## Differential Expression Analysis

### Pseudobulk Aggregation

Single-cell DE analysis requires careful statistical treatment. Simply treating
each cell as an independent replicate inflates false positives because cells from
the same donor are correlated. The **pseudobulk** approach addresses this by
summing expression counts across all cells of a given type within each biological
replicate (donor × condition), then applying bulk RNA-seq statistical methods.

Since the CELLxGENE data provides log1p-normalized values rather than raw counts,
we back-transform with `expm1()` and round to approximate integer counts. This is
a standard workaround and the pseudobulk summation over thousands of cells smooths
rounding artifacts.

```{r}
#| label: pseudobulk-aggregation
#| message: false
# Back-transform log1p to approximate counts
# Work per cell type to manage memory
cell_types_for_de <- c(
  "progenitor cell",
  "cerebral cortex neuron",
  "neuronal-restricted precursor"
)

# Create pseudobulk samples
make_pseudobulk <- function(sce, cell_type_name) {
  idx <- which(sce$cell_type == cell_type_name)
  sub <- sce[, idx]

  # Define groups
  groups <- paste(sub$condition, sub$donor_id, sep = "_")
  group_levels <- unique(groups)

  # Sum approximate counts per group
  mat <- assay(sub, "X")
  n_genes <- nrow(mat)
  pb <- matrix(0, nrow = n_genes, ncol = length(group_levels),
               dimnames = list(rownames(mat), group_levels))
  for (i in seq_along(group_levels)) {
    g <- group_levels[i]
    cells <- which(groups == g)
    if (length(cells) == 1) {
      pb[, i] <- as.numeric(round(expm1(mat[, cells])))
    } else {
      pb[, i] <- as.numeric(Matrix::rowSums(round(expm1(mat[, cells, drop = FALSE]))))
    }
  }

  # Sample metadata
  sample_info <- data.frame(
    sample    = group_levels,
    condition = sub("^(CTRL|VPA)_.*$", "\\1", group_levels),
    donor_id  = sub("^(CTRL|VPA)_", "", group_levels),
    n_cells   = as.integer(table(groups)[group_levels]),
    row.names = group_levels
  )

  list(counts = as.matrix(pb), sample_info = sample_info)
}

pseudobulk_list <- lapply(
  setNames(cell_types_for_de, cell_types_for_de),
  function(ct) make_pseudobulk(sce, ct)
)

# Show sample sizes
pb_sample_info <- do.call(rbind, lapply(names(pseudobulk_list), function(ct) {
  info <- pseudobulk_list[[ct]]$sample_info[, c("sample", "condition", "n_cells")]
  info$cell_type <- ct
  info
}))
knitr::kable(
  pb_sample_info[, c("cell_type", "sample", "condition", "n_cells")],
  caption = "Pseudobulk sample sizes",
  row.names = FALSE
)
```

### limma-voom Differential Expression

We use **limma-voom** with a paired design formula `~donor_id + condition` to
test for VPA effects while blocking on donor variability. The **empirical Bayes**
shrinkage in limma is especially valuable with small sample sizes, as it borrows
information across genes to stabilize variance estimates.

**Important caveat**: With only 2 donors per condition, statistical power is
limited. These results identify candidate genes for follow-up but should not be
treated as definitive.

```{r}
#| label: limma-voom-de
#| message: false
#| warning: false
run_limma_voom <- function(pb, cell_type_name) {
  counts <- pb$counts
  info   <- pb$sample_info

  # Create DGEList
  y <- DGEList(counts = counts, samples = info)

  # Filter lowly expressed genes
  keep <- filterByExpr(y, group = info$condition)
  y <- y[keep, , keep.lib.sizes = FALSE]

  # Normalize
  y <- calcNormFactors(y)

  # Design matrix (paired)
  design <- model.matrix(~donor_id + condition, data = info)

  # voom transformation
  v <- voom(y, design, plot = FALSE)

  # Fit
  fit <- lmFit(v, design)
  fit <- eBayes(fit)

  # Extract results for VPA effect
  # The coefficient name depends on the factor levels
  coef_name <- grep("condition", colnames(coef(fit)), value = TRUE)

  tt <- topTable(fit, coef = coef_name, number = Inf, sort.by = "none")
  tt$gene <- rownames(tt)
  tt$cell_type <- cell_type_name
  tt
}

de_results <- lapply(names(pseudobulk_list), function(ct) {
  run_limma_voom(pseudobulk_list[[ct]], ct)
})
names(de_results) <- names(pseudobulk_list)

de_all <- bind_rows(de_results)
```

### DE Summary

```{r}
#| label: de-summary
de_summary <- de_all |>
  group_by(cell_type) |>
  summarize(
    genes_tested = n(),
    up_FDR05   = sum(adj.P.Val < 0.05 & logFC > 0, na.rm = TRUE),
    down_FDR05 = sum(adj.P.Val < 0.05 & logFC < 0, na.rm = TRUE),
    up_FDR10   = sum(adj.P.Val < 0.10 & logFC > 0, na.rm = TRUE),
    down_FDR10 = sum(adj.P.Val < 0.10 & logFC < 0, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(de_summary,
              caption = "Number of DE genes per cell type (VPA vs CTRL)",
              options = list(dom = "t"))
```

### Volcano Plots

Volcano plots show the relationship between effect size (log fold change, x-axis)
and statistical significance (−log10 adjusted p-value, y-axis). Genes in the
upper corners are both large in effect and statistically significant.

```{r}
#| label: volcano-plots
#| fig-width: 16
#| fig-height: 6
#| warning: false
volcano_plots <- lapply(names(de_results), function(ct) {
  df <- de_results[[ct]]
  df$sig <- ifelse(df$adj.P.Val < 0.05, "FDR < 0.05", "NS")

  # Top genes to label
  top_genes <- df |>
    dplyr::filter(adj.P.Val < 0.05) |>
    arrange(adj.P.Val) |>
    head(15)

  ggplot(df, aes(logFC, -log10(adj.P.Val), color = sig)) +
    geom_point(size = 0.8, alpha = 0.5) +
    scale_color_manual(values = c("FDR < 0.05" = "red", "NS" = "grey60")) +
    geom_text_repel(
      data = top_genes,
      aes(label = gene),
      size = 0.9 * base_size / ggplot2::.pt,
      max.overlaps = 20, color = "black"
    ) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
    labs(
      title = ct,
      x = "log2 Fold Change (VPA vs CTRL)",
      y = "-log10(adj. P-value)",
      color = NULL
    ) +
    theme(legend.position = "bottom")
})

wrap_plots(volcano_plots, nrow = 1)
```

### Top DE Genes Table

```{r}
#| label: de-table
#| warning: false
# DT warns about data size due to embedded HTML gene links; harmless
top_de <- de_all |>
  dplyr::filter(adj.P.Val < 0.10) |>
  arrange(adj.P.Val) |>
  mutate(
    gene = gene_link(gene),
    logFC = round(logFC, 3),
    AveExpr = round(AveExpr, 2),
    t = round(t, 2),
    P.Value = signif(P.Value, 3),
    adj.P.Val = signif(adj.P.Val, 3),
    B = round(B, 2)
  ) |>
  dplyr::select(cell_type, gene, logFC, AveExpr, t, P.Value, adj.P.Val, B)

DT::datatable(
  top_de,
  caption = "DE genes (FDR < 0.10) across cell types — click gene name for GeneCards",
  escape = FALSE,
  rownames = FALSE,
  options = list(pageLength = 25, order = list(list(5, "asc")))
)
```

### Heatmap of Top DE Genes

```{r}
#| label: de-heatmap
#| fig-width: 12
#| fig-height: 10
# Get top DE genes across all cell types
top_genes_for_heatmap <- de_all |>
  dplyr::filter(adj.P.Val < 0.10) |>
  group_by(cell_type) |>
  arrange(adj.P.Val) |>
  slice_head(n = 25) |>
  ungroup() |>
  pull(gene) |>
  unique()

if (length(top_genes_for_heatmap) > 2) {
  # Build a matrix of logFC values across pseudobulk samples
  # Use the averaged log-CPM from each pseudobulk
  heatmap_data <- list()
  for (ct in names(pseudobulk_list)) {
    pb <- pseudobulk_list[[ct]]
    y <- DGEList(counts = pb$counts, samples = pb$sample_info)
    y <- calcNormFactors(y)
    lcpm <- cpm(y, log = TRUE)
    genes_avail <- intersect(top_genes_for_heatmap, rownames(lcpm))
    if (length(genes_avail) > 0) {
      mat <- lcpm[genes_avail, , drop = FALSE]
      colnames(mat) <- paste0(ct, ":", colnames(mat))
      heatmap_data[[ct]] <- mat
    }
  }

  # Combine and scale
  all_genes <- Reduce(intersect, lapply(heatmap_data, rownames))
  if (length(all_genes) > 2) {
    hm_mat <- do.call(cbind, lapply(heatmap_data, function(m) m[all_genes, ]))
    hm_scaled <- t(scale(t(hm_mat)))

    # Annotation
    ann_col <- data.frame(
      condition = sub(".*:(CTRL|VPA)_.*", "\\1", colnames(hm_mat)),
      cell_type = sub(":.*", "", colnames(hm_mat)),
      row.names = colnames(hm_mat)
    )

    pheatmap(
      hm_scaled,
      annotation_col = ann_col,
      show_rownames = (nrow(hm_scaled) <= 50),
      cluster_cols = TRUE,
      cluster_rows = TRUE,
      main = "Top DE genes across pseudobulk samples (z-scored log-CPM)",
      fontsize_row = 7
    )
  } else {
    cat("Too few shared DE genes across cell types for a meaningful heatmap.\n")
  }
} else {
  cat("Fewer than 3 DE genes at FDR < 0.10; skipping heatmap.\n")
}
```


## Gene Set Enrichment Analysis

Gene Set Enrichment Analysis (GSEA) asks whether genes belonging to known
biological pathways are systematically shifted up or down in expression. Rather
than testing one gene at a time, GSEA tests **sets of genes** that share a
function, making it more sensitive to coordinated but modest changes.

We rank genes by their limma t-statistic (which captures both effect size and
certainty) and test against **Hallmark** and **GO Biological Process** gene sets
from MSigDB.

```{r}
#| label: gsea-gene-sets
#| message: false
# Get gene sets
hallmark_df <- msigdbr(species = "Homo sapiens", collection = "H") |>
  dplyr::select(gs_name, gene_symbol)
hallmark <- split(hallmark_df$gene_symbol, hallmark_df$gs_name)

gobp_df <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP") |>
  dplyr::select(gs_name, gene_symbol)
gobp <- split(gobp_df$gene_symbol, gobp_df$gs_name)

# Combine
all_gene_sets <- c(hallmark, gobp)
```

```{r}
#| label: gsea-run
#| message: false
#| warning: false
gsea_results <- lapply(names(de_results), function(ct) {
  df <- de_results[[ct]]
  # Create ranked list by t-statistic
  ranks <- setNames(df$t, df$gene)
  ranks <- sort(ranks, decreasing = TRUE)

  # Run fgsea
  res <- fgsea(
    pathways = all_gene_sets,
    stats = ranks,
    minSize = 15,
    maxSize = 500
  )
  res$cell_type <- ct
  res
})
names(gsea_results) <- names(de_results)

gsea_all <- bind_rows(gsea_results)
```

### Top Enriched Pathways

```{r}
#| label: gsea-dotplot
#| fig-width: 10
#| fig-height: 14
#| warning: false
# Top pathways per cell type (by adjusted p-value)
top_pathways <- gsea_all |>
  dplyr::filter(padj < 0.05) |>
  group_by(cell_type) |>
  arrange(padj) |>
  slice_head(n = 15) |>
  ungroup()

if (nrow(top_pathways) > 0) {
  # Clean names for display
  top_pathways$pathway_short <- gsub("^HALLMARK_|^GOBP_", "", top_pathways$pathway)
  top_pathways$pathway_short <- gsub("_", " ", top_pathways$pathway_short)
  top_pathways$pathway_short <- tolower(top_pathways$pathway_short)

  ggplot(top_pathways, aes(
    x = NES,
    y = reorder(pathway_short, NES),
    size = -log10(padj),
    color = NES
  )) +
    geom_point() +
    scale_color_gradient2(low = "blue", mid = "grey80", high = "red", midpoint = 0) +
    facet_wrap(~cell_type, scales = "free_y", ncol = 1) +
    labs(
      title = "Top enriched pathways (GSEA, padj < 0.05)",
      x = "Normalized Enrichment Score",
      y = NULL,
      size = "-log10(padj)",
      color = "NES"
    ) +
    theme(axis.text.y = element_text(size = rel(0.6)))
} else {
  cat("No pathways reached significance at padj < 0.05.\n")
  # Show top pathways at a relaxed threshold
  top_pathways_relaxed <- gsea_all |>
    group_by(cell_type) |>
    arrange(pval) |>
    slice_head(n = 10) |>
    ungroup()

  if (nrow(top_pathways_relaxed) > 0) {
    top_pathways_relaxed$pathway_short <- gsub(
      "^HALLMARK_|^GOBP_", "", top_pathways_relaxed$pathway
    )
    top_pathways_relaxed$pathway_short <- gsub("_", " ", top_pathways_relaxed$pathway_short)
    top_pathways_relaxed$pathway_short <- tolower(top_pathways_relaxed$pathway_short)

    ggplot(top_pathways_relaxed, aes(
      x = NES,
      y = reorder(pathway_short, NES),
      size = -log10(pval),
      color = NES
    )) +
      geom_point() +
      scale_color_gradient2(low = "blue", mid = "grey80", high = "red", midpoint = 0) +
      facet_wrap(~cell_type, scales = "free_y", ncol = 1) +
      labs(
        title = "Top pathways by nominal p-value (none reached padj < 0.05)",
        x = "Normalized Enrichment Score",
        y = NULL,
        size = "-log10(p-value)",
        color = "NES"
      ) +
      theme(axis.text.y = element_text(size = rel(0.6)))
  }
}
```

### GSEA Results Table

```{r}
#| label: gsea-table
gsea_table <- gsea_all |>
  dplyr::filter(pval < 0.05) |>
  arrange(pval) |>
  mutate(
    NES = round(NES, 2),
    pval = signif(pval, 3),
    padj = signif(padj, 3),
    size = as.integer(size)
  ) |>
  dplyr::select(cell_type, pathway, NES, pval, padj, size) |>
  head(100)

DT::datatable(
  gsea_table,
  caption = "GSEA results (nominal p < 0.05)",
  options = list(pageLength = 25, order = list(list(3, "asc")))
)
```


## Known Biology Check

We examine genes known to be responsive to VPA or involved in neural development
to see whether their expression patterns are consistent with expectations.

**Selected genes**:

- **HDAC targets / chromatin remodeling**: *HDAC1*, *HDAC2*, *HDAC3*, *EP300*, *CREBBP*
- **Neural progenitor markers**: *PAX6*, *SOX2*, *NES*, *HES1*, *NOTCH1*
- **Neuronal differentiation**: *DCX*, *TBR1*, *NEUROD1*, *NEUROD2*, *MAP2*
- **Cortical layer markers**: *SATB2*, *BCL11B* (CTIP2), *TLE4*
- **Apoptosis / cell stress**: *TP53*, *BCL2*, *BAX*

```{r}
#| label: known-genes-data
known_genes <- c(
  "HDAC1", "HDAC2", "HDAC3", "EP300", "CREBBP",
  "PAX6", "SOX2", "NES", "HES1", "NOTCH1",
  "DCX", "TBR1", "NEUROD1", "NEUROD2", "MAP2",
  "SATB2", "BCL11B", "TLE4",
  "TP53", "BCL2", "BAX"
)

# Filter to genes present in data
known_genes_present <- known_genes[known_genes %in% rownames(sce)]
cat("Genes present in dataset:", length(known_genes_present), "of", length(known_genes), "\n")
cat("Missing:", setdiff(known_genes, known_genes_present), "\n")
```

### Violin Plots

```{r}
#| label: known-genes-violin
#| fig-width: 16
#| fig-height: 10
#| warning: false
#| results: asis
# Extract expression for known genes
if (length(known_genes_present) > 0) {
  expr_mat <- as.matrix(assay(sce, "X")[known_genes_present, , drop = FALSE])
  expr_df <- as.data.frame(t(expr_mat))
  expr_df$condition <- sce$condition
  expr_df$cell_type <- sce$cell_type

  expr_long <- expr_df |>
    tidyr::pivot_longer(
      cols = all_of(known_genes_present),
      names_to = "gene",
      values_to = "expression"
    )

  # Only plot for the 3 major cell types
  expr_long_major <- expr_long |>
    dplyr::filter(cell_type %in% cell_types_for_de)

  # Split into groups of 6 genes to keep panels readable
  genes_per_plot <- 6
  gene_groups <- split(
    unique(expr_long_major$gene),
    ceiling(seq_along(unique(expr_long_major$gene)) / genes_per_plot)
  )

  for (i in seq_along(gene_groups)) {
    p <- expr_long_major |>
      dplyr::filter(gene %in% gene_groups[[i]]) |>
      ggplot(aes(condition, expression, fill = condition)) +
      geom_violin(scale = "width", alpha = 0.7, linewidth = 0.3) +
      scale_fill_manual(values = condition_colors) +
      facet_grid(gene ~ cell_type, scales = "free_y", switch = "y") +
      theme(
        strip.placement = "outside",
        strip.text.y.left = element_text(angle = 0, size = rel(0.65)),
        strip.text.x = element_text(size = rel(0.7)),
        axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.6))
      ) +
      labs(
        title = paste0(
          "VPA-response and neural genes (group ", i,
          " of ", length(gene_groups), ")"
        ),
        x = NULL, y = "log1p expression", fill = "Condition"
      )
    print(p)
  }
}
```

### Feature Plots (UMAP) for Top DE Genes

```{r}
#| label: feature-plots
#| fig-width: 16
#| fig-height: 12
#| warning: false
# Pick top DE genes across cell types
top_de_genes <- de_all |>
  dplyr::filter(adj.P.Val < 0.10) |>
  arrange(adj.P.Val) |>
  pull(gene) |>
  unique() |>
  head(8)

# If too few DE genes, use top by nominal p
if (length(top_de_genes) < 4) {
  top_de_genes <- de_all |>
    arrange(P.Value) |>
    pull(gene) |>
    unique() |>
    head(8)
}

if (length(top_de_genes) > 0) {
  # Extract expression
  expr_top <- as.matrix(assay(sce, "X")[top_de_genes, , drop = FALSE])
  umap_expr <- umap_df
  for (g in top_de_genes) {
    umap_expr[[g]] <- expr_top[g, ]
  }

  feature_plots <- lapply(top_de_genes, function(g) {
    ggplot(umap_expr[sample(nrow(umap_expr)), ], aes(UMAP1, UMAP2, color = .data[[g]])) +
      geom_point(size = 0.2, alpha = 0.5) +
      scale_color_viridis(option = "magma") +
      labs(title = g, color = "log1p\nexpr") +
      theme(legend.key.width = unit(0.5, "cm"))
  })

  wrap_plots(feature_plots, ncol = 4)
}
```


## Draft Results

```{r}
#| label: results-data
#| echo: false
n_cells <- ncol(sce)
n_ctrl <- sum(sce$condition == "CTRL")
n_vpa <- sum(sce$condition == "VPA")
n_donors <- length(unique(sce$donor_id))
n_cell_types <- length(unique(sce$cell_type))

# Composition changes
comp_summary <- comp_cond |>
  dplyr::select(condition, cell_type, proportion) |>
  tidyr::pivot_wider(names_from = condition, values_from = proportion) |>
  mutate(change_pct = round((VPA - CTRL) * 100, 1))

# DE counts
n_de_005 <- sum(de_all$adj.P.Val < 0.05, na.rm = TRUE)
n_de_010 <- sum(de_all$adj.P.Val < 0.10, na.rm = TRUE)

# Top pathway
top_pw <- gsea_all |> arrange(pval) |> head(1)
```

### Results Summary

We analyzed `r format(n_cells, big.mark = ",")` cells from
`r n_donors` donors across `r n_cell_types` cell types, comparing CTRL
(`r format(n_ctrl, big.mark = ",")` cells) to VPA-treated
(`r format(n_vpa, big.mark = ",")` cells) human dorsal forebrain organoids.

**Cell type composition**: VPA treatment was associated with notable shifts in
cell type proportions:

```{r}
#| label: results-comp-table
#| echo: false
DT::datatable(
  comp_summary |>
    mutate(
      CTRL = paste0(round(CTRL * 100, 1), "%"),
      VPA = paste0(round(VPA * 100, 1), "%"),
      `
Change(pp)
` = change_pct
    ) |>
    dplyr::select(cell_type, CTRL, VPA, `
Change(pp)
`),
  caption = "Cell type proportion changes (percentage points)",
  options = list(dom = "t"),
  rownames = FALSE
)
```

**Differential expression**: Across the three major cell types tested,
we identified `r n_de_005` genes at FDR < 0.05 and `r n_de_010` genes at
FDR < 0.10.

**Pathway enrichment**: The top enriched pathway was
*`r gsub("_", " ", top_pw$pathway[1])`* in `r top_pw$cell_type[1]`
(NES = `r round(top_pw$NES[1], 2)`, nominal p = `r signif(top_pw$pval[1], 3)`).

**Limitations**: With only `r n_donors` donors per condition, these results
are exploratory. The use of back-transformed log1p data rather than raw counts
introduces approximation, though pseudobulk aggregation mitigates this.


## Draft Methods

### Data Source

Single-cell RNA-seq data were obtained from CELLxGENE
(dataset ID: `3a8aec06-3309-4d37-b75d-c59f5f6d55a6`), originating from
Yentür et al. (2025, bioRxiv). The dataset comprises human dorsal forebrain
organoids from `r n_donors` donors under control and VPA-treated conditions
(`r format(n_cells, big.mark = ",")` cells total).

### Preprocessing

Data were provided as log1p-normalized expression values. UMAP coordinates and
cell type annotations from the original authors were used without modification.

### Differential Expression

Pseudobulk expression profiles were generated by summing back-transformed
(expm1) approximate counts per gene across cells within each donor × condition
× cell type group. Differential expression was performed using the limma-voom
pipeline (Ritchie et al. 2015) with a paired design formula (~donor_id +
condition). Genes were filtered using `filterByExpr()`, library sizes
normalized with TMM (`calcNormFactors()`), and precision weights estimated
via `voom()`. Empirical Bayes moderation was applied with `eBayes()`.
P-values were adjusted for multiple testing using the Benjamini-Hochberg method.

### Pathway Analysis

Gene set enrichment analysis was performed using fgsea (Korotkevich et al. 2021)
with genes ranked by limma t-statistic. MSigDB Hallmark and GO Biological Process
gene sets were tested (gene set sizes 15–500).

### Software

```{r}
#| label: methods-software
#| echo: false
#| results: asis
pkgs <- c("R", "SingleCellExperiment", "edgeR", "limma", "fgsea", "msigdbr",
           "ggplot2", "plotly")
versions <- sapply(pkgs, function(p) {
  if (p == "R") {
    paste(R.version$major, R.version$minor, sep = ".")
  } else {
    as.character(packageVersion(p))
  }
})
cat(paste(
  paste0("**", pkgs, "** v", versions),
  collapse = ", "
), "\n")
```


## References

- Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W, Smyth GK (2015).
  "limma powers differential expression analyses for RNA-sequencing and
  microarray studies." *Nucleic Acids Research* 43(7):e47.
  <https://doi.org/10.1093/nar/gkv007>

- Law CW, Chen Y, Shi W, Smyth GK (2014). "voom: precision weights
  unlock linear model analysis tools for RNA-seq read counts." *Genome
  Biology* 15:R29. <https://doi.org/10.1186/gb-2014-15-2-r29>

- Robinson MD, McCarthy DJ, Smyth GK (2010). "edgeR: a Bioconductor
  package for differential expression analysis of digital gene expression
  data." *Bioinformatics* 26(1):139–140.
  <https://doi.org/10.1093/bioinformatics/btp616>

- Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A
  (2021). "Fast gene set enrichment analysis." *bioRxiv*.
  <https://doi.org/10.1101/060012>

- Liberzon A, Birger C, Thorvaldsdóttir H, Ghandi M, Mesirov JP, Tamayo P
  (2015). "The Molecular Signatures Database Hallmark Gene Set Collection."
  *Cell Systems* 1(6):417–425.
  <https://doi.org/10.1016/j.cels.2015.12.004>

- Amezquita RA, Lun ATL, Becht E, Carey VJ, Carpp LN, Geistlinger L,
  Marini F, Rue-Albrecht K, Risso D, Soneson C, Waldron L, Pagès H,
  Smith ML, Huber W, Morgan M, Gottardo R, Hicks SC (2020).
  "Orchestrating single-cell analysis with Bioconductor." *Nature Methods*
  17:137–145. <https://doi.org/10.1038/s41592-019-0654-x>

- McInnes L, Healy J, Melville J (2018). "UMAP: Uniform Manifold
  Approximation and Projection for Dimension Reduction."
  *arXiv:1802.03426*. <https://doi.org/10.48550/arXiv.1802.03426>

- Squair JW, Gautier M, Kathe C, Anderson MA, James ND, Hutson TH,
  Hudelle R, Qaber T, Matson KJE, Baez-Nieto D, Levine AJ, Bhatt DK,
  Courtine G (2021). "Confronting false discoveries in single-cell
  differential expression." *Nature Communications* 12:5692.
  <https://doi.org/10.1038/s41467-021-25960-2>


::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info
#| echo: false
#| comment: ""
sessioninfo::session_info(info = "all")
`
``
:::
