---
title: "Control vs VPA-Treated Dorsal Forebrain Organoids: scRNA-seq Analysis"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: last-modified
date-format: "YYYY-MM-DD HH:mm"
format:
  html:
    page-layout: full
    embed-resources: true
    toc: true
    code-fold: true
    code-summary: "Show code"
    include-in-header:
      text: |
        <style>
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
        </style>
---

## Introduction

Valproic acid (VPA) is a widely prescribed antiepileptic drug that also acts as a
**histone deacetylase (HDAC) inhibitor**. Prenatal exposure to VPA is an established
teratogen and has been linked to increased risk of autism spectrum disorder and other
neurodevelopmental conditions.

This analysis examines single-cell RNA-seq data from
[Yentür et al. (2025, bioRxiv)](https://doi.org/10.1101/2025.06.24.661272),
comparing **control** vs **VPA-treated** human dorsal forebrain organoids.
The dataset was obtained from
[CELLxGENE](https://cellxgene.cziscience.com/) and contains 26,499 cells across
9 cell types annotated by the original authors (RG, IPC/Newborn ExN, Young dExN,
Mature dExN, InhN, Cajal-Retzius N, RG stressed, dExN stressed, Unknown).

**Experimental design**: 2 donors, each contributing organoids to both control and
VPA conditions (a paired design). Because we have only 2 biological replicates per
condition, all differential expression results should be considered **exploratory**.

**What this analysis covers**:

1. Visualization of cell populations (UMAP)
2. Cell type composition changes under VPA
3. Pseudobulk differential expression per cell type (limma-voom, paired design)
4. Gene set enrichment analysis (GSEA with Hallmark and GO gene sets)
5. Expression of known VPA-response and neural development genes


## Setup & Data Loading

```{r}
#| label: setup
#| message: false
#| warning: false
library(SingleCellExperiment)
library(SummarizedExperiment)
library(Seurat)
library(edgeR)
library(limma)
library(ggplot2)
library(plotly)
library(patchwork)
library(DT)
library(dplyr)
library(tidyr)
library(ggrepel)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(pheatmap)
library(scales)
library(glue)
library(here)
library(Matrix)
library(viridis)

theme_set(theme_bw(base_size = 14))
```

```{r}
#| label: load-data
sce <- readRDS(here::here("analysis", "vpa_organoids_sce.rds"))

# Map Ensembl IDs to gene symbols where available
# feature_name contains gene symbols for annotated genes, ENSG IDs otherwise
gene_symbols <- as.character(rowData(sce)$feature_name)
is_symbol <- !grepl("^ENSG", gene_symbols)

# Make symbols unique (some may be duplicated)
symbol_map <- gene_symbols
symbol_map[is_symbol] <- make.unique(symbol_map[is_symbol])
names(symbol_map) <- rownames(sce)

# Create a lookup from ENSG to symbol
ensg_to_symbol <- setNames(gene_symbols, rownames(sce))

# Rename rows to gene symbols where available
rownames(sce) <- symbol_map
```

```{r}
#| label: convert-to-seurat
#| message: false

# The CELLxGENE data provides log1p-normalized values (no raw counts).
# We back-transform to approximate counts for Seurat's counts layer and
# store the original log1p-normalized values as the data layer.
norm_mat <- assay(sce, "X")
norm_sparse <- as(norm_mat, "CsparseMatrix")
approx_counts <- round(expm1(norm_sparse))

obj <- CreateSeuratObject(
  counts = approx_counts,
  meta.data = as.data.frame(colData(sce))
)

# Store original log1p-normalized data as the data layer
obj[["RNA"]]$data <- norm_sparse

# Store the authors' pre-computed UMAP as a named reduction
umap_coords <- reducedDim(sce, "X_umap")
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")
rownames(umap_coords) <- colnames(obj)
obj[["author_umap"]] <- CreateDimReducObject(
  embeddings = umap_coords, key = "authorUMAP_", assay = "RNA"
)

# Free memory
rm(sce, norm_mat, norm_sparse, approx_counts, umap_coords)
gc()
```

The dataset contains **`r format(ncol(obj), big.mark = ",")`
cells** and **`r format(nrow(obj), big.mark = ",")`
genes** (`r sum(is_symbol)` with gene symbols, `r sum(!is_symbol)` Ensembl IDs only).

```{r}
#| label: color-palettes
# Use the authors' original cell type annotations (author_cell_type column)
# rather than the coarser CELLxGENE Cell Ontology mapping (cell_type column).
# The CELLxGENE cell_type column collapses 9 types into 5 broad categories.
cell_type_colors <- c(
  "RG"               = "#1b9e77",
  "IPC/Newborn ExN"  = "#d95f02",
  "Young dExN"       = "#7570b3",
  "Mature dExN"      = "#e7298a",
  "InhN"             = "#66a61e",
  "Cajal-Retzius N"  = "#e6ab02",
  "RG stressed"      = "#a6d854",
  "dExN stressed"    = "#b3b3b3",
  "Unknown"          = "#999999"
)

condition_colors <- c(
  "CTRL" = "#4393c3",
  "VPA"  = "#d6604d"
)

donor_colors <- c(
  "#66c2a5",
  "#fc8d62"
)
names(donor_colors) <- sort(unique(obj$donor_id))
```

```{r}
#| label: gene-link-helper
#' Create a clickable GeneCards link for a gene symbol
gene_link <- function(symbol) {
  paste0(
    '<a href="https://www.genecards.org/cgi-bin/carddisp.pl?gene=',
    symbol, '" target="_blank">', symbol, '</a>'
  )
}
```


## Experimental Design Overview

The experiment uses a **paired design**: each donor contributes cells to both
conditions. This lets us account for donor-to-donor variability when testing
for VPA effects.

```{r}
#| label: design-tables
knitr::kable(
  as.data.frame.matrix(table(obj$condition, obj$donor_id)),
  caption = "Cells per condition and donor"
)

knitr::kable(
  as.data.frame.matrix(table(obj$condition, obj$author_cell_type)),
  caption = "Cells per condition and cell type"
)
```

```{r}
#| label: design-table-dt
design_df <- as.data.frame(
  table(Condition = obj$condition, Donor = obj$donor_id, CellType = obj$author_cell_type)
) |>
  dplyr::filter(Freq > 0) |>
  tidyr::pivot_wider(names_from = Condition, values_from = Freq, values_fill = 0)

DT::datatable(design_df, caption = "Cell counts by donor, condition, and cell type",
              options = list(pageLength = 20))
```


## UMAP Visualizations

UMAP (Uniform Manifold Approximation and Projection) is a dimensionality reduction
technique that projects high-dimensional gene expression data into 2D for
visualization. Cells that are close together on the UMAP have similar expression
profiles. The UMAP coordinates here were computed by the original study authors.

```{r}
#| label: umap-data
# Build a data frame from the Seurat object for custom plots and Plotly
umap_df <- data.frame(
  Embeddings(obj, reduction = "author_umap"),
  cell_type  = obj$author_cell_type,
  condition  = obj$condition,
  donor_id   = obj$donor_id,
  cell_phase = obj$cell_phase
)
colnames(umap_df)[1:2] <- c("UMAP1", "UMAP2")
```

### UMAP by Cell Type

```{r}
#| label: umap-cell-type
#| fig-width: 9
#| fig-height: 7
#| fig-alt: "UMAP scatter plot of 26,499 cells colored by 9 author-annotated cell types. Major clusters include radial glia, excitatory neurons at various maturation stages, inhibitory neurons, and stressed subtypes."
DimPlot(obj, reduction = "author_umap", group.by = "author_cell_type",
        cols = cell_type_colors, pt.size = 0.3, alpha = 0.5) +
  labs(title = "UMAP colored by cell type", color = "Cell type")
```

### UMAP by Condition

```{r}
#| label: umap-condition
#| fig-width: 9
#| fig-height: 7
#| fig-alt: "UMAP scatter plot colored by treatment condition (CTRL blue, VPA red). Cells from both conditions are intermixed across most clusters."
DimPlot(obj, reduction = "author_umap", group.by = "condition",
        cols = condition_colors, pt.size = 0.3, alpha = 0.4) +
  labs(title = "UMAP colored by condition", color = "Condition")
```

### UMAP Split by Condition

```{r}
#| label: umap-split-condition
#| fig-width: 14
#| fig-height: 6
#| fig-alt: "Side-by-side UMAP plots for CTRL and VPA conditions, colored by cell type. Allows visual comparison of cluster occupancy between conditions."
DimPlot(obj, reduction = "author_umap", group.by = "author_cell_type",
        split.by = "condition", cols = cell_type_colors, pt.size = 0.3, alpha = 0.5) +
  labs(title = "UMAP split by condition", color = "Cell type")
```

### UMAP by Donor

```{r}
#| label: umap-donor
#| fig-width: 9
#| fig-height: 7
#| fig-alt: "UMAP scatter plot colored by donor identity. Cells from both donors are intermixed across clusters, indicating no strong donor-specific batch effect."
DimPlot(obj, reduction = "author_umap", group.by = "donor_id",
        cols = donor_colors, pt.size = 0.3, alpha = 0.4) +
  labs(title = "UMAP colored by donor", color = "Donor")
```

### UMAP by Cell Cycle Phase

```{r}
#| label: umap-cell-phase
#| fig-width: 9
#| fig-height: 7
#| fig-alt: "UMAP scatter plot colored by cell cycle phase (G1, S, G2M). Cycling cells are concentrated in progenitor regions."
DimPlot(obj, reduction = "author_umap", group.by = "cell_phase",
        pt.size = 0.3, alpha = 0.4) +
  labs(title = "UMAP colored by cell cycle phase", color = "Cell phase")
```

### Interactive UMAP (Plotly)

:::{.column-screen}
```{r}
#| label: umap-plotly
#| fig-width: 10
#| fig-height: 7
# Subsample for plotly performance
set.seed(42)
idx <- sample(nrow(umap_df), min(10000, nrow(umap_df)))
umap_sub <- umap_df[idx, ]

p <- plot_ly(
  umap_sub,
  x = ~UMAP1, y = ~UMAP2,
  color = ~cell_type,
  colors = cell_type_colors,
  text = ~paste("Cell type:", cell_type,
                "<br>Condition:", condition,
                "<br>Donor:", donor_id),
  type = "scatter", mode = "markers",
  marker = list(size = 3, opacity = 0.6),
  width = 900, height = 650
) |>
  layout(
    title = "Interactive UMAP (10,000 cell subsample)",
    xaxis = list(title = "UMAP1"),
    yaxis = list(title = "UMAP2")
  )
p
```
:::


## De Novo Clustering and Cell Type Annotation

The cell type labels above come from the original study authors. Can we recover
similar groupings independently? Below we cluster cells from scratch using
Seurat's standard pipeline (HVG selection → PCA → UMAP → graph-based Louvain
clustering), then annotate clusters with canonical brain marker genes and compare
to the authors' annotations.

### Feature Selection and Dimensionality Reduction

```{r}
#| label: denovo-hvg-pca
#| message: false
#| warning: false

# Select the top 2,000 highly variable genes using Seurat's vst method
obj <- FindVariableFeatures(obj, nfeatures = 2000)
cat(length(VariableFeatures(obj)), "highly variable genes selected\n")

# Scale data (center and scale each gene across cells)
obj <- ScaleData(obj, verbose = FALSE)

# PCA on the variable features
set.seed(42)
obj <- RunPCA(obj, npcs = 30, verbose = FALSE)
```

Highly variable genes (HVGs) are genes whose expression varies more across cells
than expected given their average expression level. These carry the biological
signal that distinguishes cell types. Seurat's `FindVariableFeatures` uses
variance-stabilizing transformation to select the top 2,000 HVGs.

```{r}
#| label: denovo-elbow
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Elbow plot showing variance explained by each of 30 principal components. A clear elbow appears around PC 10-15, with diminishing returns beyond PC 20."
ElbowPlot(obj, ndims = 30) +
  labs(title = "PCA elbow plot (de novo HVGs)")
```

PCA (Principal Component Analysis) compresses the 2,000-gene expression matrix
into a small number of components that capture the dominant axes of variation.
The "elbow plot" shows how much variance each component explains — we use the
first 20 PCs (before the curve flattens) for downstream steps.

```{r}
#| label: denovo-umap
#| message: false
#| warning: false

# UMAP from PCA embeddings (use uwot/cosine, the Seurat v5 default)
set.seed(42)
obj <- RunUMAP(obj, dims = 1:20, seed.use = 42,
               umap.method = "uwot", metric = "cosine",
               reduction.name = "denovo_umap", reduction.key = "denovoUMAP_",
               verbose = FALSE)
```

### Graph-Based Clustering

```{r}
#| label: denovo-cluster
#| message: false

# Build a shared-nearest-neighbour graph and cluster with Louvain
obj <- FindNeighbors(obj, dims = 1:20, verbose = FALSE)
obj <- FindClusters(obj, resolution = 0.5, verbose = FALSE)

cluster_ids <- Idents(obj)
cat("Louvain detected", length(levels(cluster_ids)), "clusters\n")

knitr::kable(
  data.frame(
    Cluster = names(sort(table(cluster_ids), decreasing = TRUE)),
    Cells = as.integer(sort(table(cluster_ids), decreasing = TRUE))
  ),
  row.names = FALSE,
  caption = "Cluster sizes"
)
```

Graph-based clustering connects each cell to its *k* nearest neighbours in PCA
space, forming a shared-nearest-neighbour (SNN) graph weighted by Jaccard
similarity. The **Louvain algorithm** then detects communities — groups of cells
that are more densely connected to each other than to the rest of the network.

### Cell Type Annotation via Marker Genes

```{r}
#| label: denovo-markers
#| fig-width: 10
#| fig-height: 6
#| message: false
#| fig-alt: "Heatmap of mean marker gene expression per Louvain cluster. Rows are canonical cell type marker sets and columns are cluster IDs. Color intensity indicates expression level."

# Canonical brain cell type marker genes
marker_sets <- list(
  "RG"               = c("PAX6", "SOX2", "HES1", "NES", "VIM"),
  "IPC/Newborn ExN"  = c("EOMES", "NEUROG2", "NEUROD1"),
  "Young dExN"       = c("DCX", "NEUROD2", "TBR1"),
  "Mature dExN"      = c("SATB2", "SLC17A7", "BCL11B"),
  "InhN"             = c("GAD1", "GAD2", "DLX2", "DLX5"),
  "Cajal-Retzius N"  = c("RELN", "CALB2", "TP73")
)

# Compute mean expression of each marker set per cluster using AverageExpression
all_markers <- unlist(marker_sets)
all_markers <- all_markers[all_markers %in% rownames(obj)]

avg_expr <- AverageExpression(obj, features = all_markers,
                               group.by = "seurat_clusters",
                               layer = "data")
avg_mat <- avg_expr$RNA

# Build a score matrix: mean marker expression per cell type per cluster
clusters <- levels(Idents(obj))
score_mat <- matrix(0, nrow = length(marker_sets), ncol = length(clusters),
                    dimnames = list(names(marker_sets), clusters))

for (ct_name in names(marker_sets)) {
  markers <- marker_sets[[ct_name]]
  markers <- markers[markers %in% rownames(avg_mat)]
  if (length(markers) == 0) next
  if (length(markers) == 1) {
    score_mat[ct_name, ] <- avg_mat[markers, ]
  } else {
    score_mat[ct_name, ] <- colMeans(avg_mat[markers, , drop = FALSE])
  }
}

# Heatmap of marker scores
pheatmap(score_mat,
         cluster_rows = FALSE,
         main = "Mean marker expression per cluster",
         color = viridis(100),
         angle_col = 0)
```

These marker genes are well established in developmental neuroscience:

- **Radial glia (RG)**: PAX6, SOX2, HES1, NES, VIM — neural stem/progenitor markers
- **Intermediate progenitors / Newborn excitatory neurons (IPC/Newborn ExN)**:
  EOMES (TBR2), NEUROG2, NEUROD1
- **Young deep-layer excitatory neurons (Young dExN)**: DCX, NEUROD2, TBR1
- **Mature deep-layer excitatory neurons (Mature dExN)**: SATB2, SLC17A7, BCL11B (CTIP2)
- **Inhibitory neurons (InhN)**: GAD1, GAD2, DLX2, DLX5
- **Cajal-Retzius neurons**: RELN, CALB2, TP73

```{r}
#| label: denovo-assign-types

# Assign each cluster to the cell type with the highest marker score
best_type <- apply(score_mat, 2, function(x) names(which.max(x)))
obj$denovo_cell_type <- unname(best_type[as.character(Idents(obj))])

type_assignment <- data.frame(
  Cluster     = clusters,
  `Cell type` = best_type[clusters],
  Cells       = as.integer(table(Idents(obj))[clusters]),
  check.names = FALSE
)
knitr::kable(type_assignment, row.names = FALSE, caption = "De novo cell type assignments")
```


### Comparing to Author Annotations

```{r}
#| label: denovo-comparison-umap
#| fig-width: 14
#| fig-height: 12
#| fig-alt: "Four-panel UMAP comparison. Top row shows the authors' UMAP with author labels (left) and de novo labels (right). Bottom row shows the de novo UMAP with author labels (left) and de novo labels (right). Most cell types show strong agreement between the two annotation schemes."

# Color palette for de novo types (reuse author palette where possible)
denovo_types <- sort(unique(obj$denovo_cell_type))
denovo_colors <- cell_type_colors[denovo_types]
# Fill any missing with defaults
missing <- is.na(denovo_colors)
if (any(missing)) {
  denovo_colors[missing] <- scales::hue_pal()(sum(missing))
}

# 4-panel comparison using DimPlot
p1 <- DimPlot(obj, reduction = "author_umap", group.by = "author_cell_type",
              cols = cell_type_colors, pt.size = 0.2, alpha = 0.4) +
  labs(title = "Authors' UMAP, authors' labels", color = "Cell type") +
  theme(legend.position = "bottom")

p2 <- DimPlot(obj, reduction = "author_umap", group.by = "denovo_cell_type",
              cols = denovo_colors, pt.size = 0.2, alpha = 0.4) +
  labs(title = "Authors' UMAP, de novo labels", color = "De novo type") +
  theme(legend.position = "bottom")

p3 <- DimPlot(obj, reduction = "denovo_umap", group.by = "author_cell_type",
              cols = cell_type_colors, pt.size = 0.2, alpha = 0.4) +
  labs(title = "De novo UMAP, authors' labels", color = "Cell type") +
  theme(legend.position = "bottom")

p4 <- DimPlot(obj, reduction = "denovo_umap", group.by = "denovo_cell_type",
              cols = denovo_colors, pt.size = 0.2, alpha = 0.4) +
  labs(title = "De novo UMAP, de novo labels", color = "De novo type") +
  theme(legend.position = "bottom")

(p1 + p2) / (p3 + p4)
```

```{r}
#| label: denovo-concordance

# --- Cross-tabulation ---
cross_tab <- table(
  Author  = obj$author_cell_type,
  DeNovo  = obj$denovo_cell_type
)

DT::datatable(
  as.data.frame.matrix(cross_tab),
  caption = "Cross-tabulation: author cell types (rows) vs de novo types (columns)",
  options = list(dom = "t", pageLength = 20)
)
```

```{r}
#| label: denovo-confusion-heatmap
#| fig-width: 9
#| fig-height: 7
#| fig-alt: "Row-normalized confusion matrix heatmap comparing author cell type annotations (rows) to de novo cluster assignments (columns). Values show the proportion of each author type assigned to each de novo type."

# Row-normalized confusion matrix (proportion of each author type assigned
# to each de novo type)
cross_prop <- sweep(cross_tab, 1, rowSums(cross_tab), "/")

pheatmap(cross_prop,
         cluster_rows = FALSE, cluster_cols = FALSE,
         main = "Confusion matrix (row-normalized)",
         color = viridis(100),
         display_numbers = TRUE,
         number_format = "%.2f",
         angle_col = 45)
```

```{r}
#| label: denovo-ari

# --- Adjusted Rand Index (computed from contingency table) ---
# ARI = (Index - Expected) / (Max - Expected)
compute_ari <- function(tab) {
  a <- sum(choose(tab, 2))
  row_sums <- rowSums(tab)
  col_sums <- colSums(tab)
  b <- sum(choose(row_sums, 2))
  c_val <- sum(choose(col_sums, 2))
  n <- sum(tab)
  d <- choose(n, 2)
  expected <- b * c_val / d
  max_index <- 0.5 * (b + c_val)
  if (max_index == expected) return(1)
  (a - expected) / (max_index - expected)
}

ari <- compute_ari(cross_tab)
```

**Adjusted Rand Index: `r round(ari, 3)`**

The ARI measures agreement between two clusterings, adjusted for chance
(0 = random, 1 = perfect agreement). Values above 0.5 indicate substantial
agreement.

::: {.callout-note}
## Why "stressed" cell types cannot be recovered de novo

The original authors identified "RG stressed" and "dExN stressed" subtypes using
the [**Gruffi**](https://github.com/jn-goe/gruffi) algorithm, which detects
cells enriched for stress/apoptosis gene signatures that arise from tissue
dissociation or organoid culture artifacts. These are not distinct developmental
lineages — they are subsets of RG and dExN cells with elevated stress gene
expression.

Because our marker-based annotation uses developmental identity markers (PAX6,
DCX, etc.) rather than stress signatures, de novo clustering correctly merges
these cells back into their parent types (RG, Young/Mature dExN). This is
expected and does not indicate a failure of the clustering.
:::


## Cell Type Composition Analysis

A key question in case-control scRNA-seq studies is whether the **proportions** of
cell types shift between conditions. Changes in composition can reflect altered
differentiation, proliferation, or survival.

```{r}
#| label: composition-data
comp_df <- as.data.frame(table(
  condition = obj$condition,
  cell_type = obj$author_cell_type,
  donor_id  = obj$donor_id
))

# Compute proportions within each condition × donor
comp_df <- comp_df |>
  group_by(condition, donor_id) |>
  mutate(
    total = sum(Freq),
    proportion = Freq / total
  ) |>
  ungroup()
```

### Stacked Bar Chart

```{r}
#| label: composition-stacked
#| fig-width: 10
#| fig-height: 6
#| fig-alt: "Stacked bar chart comparing cell type proportions between CTRL and VPA conditions, showing shifts in the relative abundance of each cell type."
# Per condition (aggregated across donors)
comp_cond <- comp_df |>
  group_by(condition, cell_type) |>
  summarize(Freq = sum(Freq), .groups = "drop") |>
  group_by(condition) |>
  mutate(proportion = Freq / sum(Freq)) |>
  ungroup()

ggplot(comp_cond, aes(condition, proportion, fill = cell_type)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = cell_type_colors) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cell type composition by condition",
    x = "Condition", y = "Proportion", fill = "Cell type"
  )
```

### Composition by Donor

```{r}
#| label: composition-by-donor
#| fig-width: 12
#| fig-height: 6
#| fig-alt: "Stacked bar chart showing cell type proportions for each donor-condition combination, revealing donor-level variation in composition."
comp_df$sample <- paste(comp_df$condition, comp_df$donor_id, sep = "\n")

ggplot(comp_df, aes(sample, proportion, fill = cell_type)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = cell_type_colors) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cell type composition by condition and donor",
    x = "Sample", y = "Proportion", fill = "Cell type"
  )
```

### Composition Table

```{r}
#| label: composition-table
comp_table <- comp_cond |>
  mutate(
    label = paste0(Freq, " (", round(proportion * 100, 1), "%)")
  ) |>
  dplyr::select(condition, cell_type, label) |>
  tidyr::pivot_wider(names_from = condition, values_from = label)

DT::datatable(comp_table, caption = "Cell counts and proportions per condition",
              escape = FALSE, options = list(dom = "t"))
```


## Differential Expression Analysis

### Pseudobulk Aggregation

Single-cell DE analysis requires careful statistical treatment. Simply treating
each cell as an independent replicate inflates false positives because cells from
the same donor are correlated. The **pseudobulk** approach addresses this by
summing expression counts across all cells of a given type within each biological
replicate (donor × condition), then applying bulk RNA-seq statistical methods.

The CELLxGENE data provides log1p-normalized values rather than raw counts.
Approximate integer counts were already computed during Seurat object creation
(back-transforming with `expm1()` and rounding). The pseudobulk step sums these
approximate counts per group; summation over thousands of cells smooths rounding
artifacts.

```{r}
#| label: pseudobulk-aggregation
#| message: false
# Sum approximate counts per cell type × donor × condition for pseudobulk DE
cell_types_for_de <- c(
  "RG",
  "IPC/Newborn ExN",
  "Young dExN",
  "Mature dExN",
  "InhN"
)

# Create pseudobulk samples from the Seurat counts layer
# (already back-transformed approximate counts)
make_pseudobulk <- function(obj, cell_type_name) {
  idx <- which(obj$author_cell_type == cell_type_name)
  sub <- obj[, idx]

  # Define groups
  groups <- paste(sub$condition, sub$donor_id, sep = "_")
  group_levels <- unique(groups)

  # Sum counts per group (counts layer already has approximate integers)
  mat <- GetAssayData(sub, layer = "counts")
  n_genes <- nrow(mat)
  pb <- matrix(0, nrow = n_genes, ncol = length(group_levels),
               dimnames = list(rownames(mat), group_levels))
  for (i in seq_along(group_levels)) {
    g <- group_levels[i]
    cells <- which(groups == g)
    if (length(cells) == 1) {
      pb[, i] <- as.numeric(mat[, cells])
    } else {
      pb[, i] <- as.numeric(Matrix::rowSums(mat[, cells, drop = FALSE]))
    }
  }

  # Sample metadata
  sample_info <- data.frame(
    sample    = group_levels,
    condition = sub("^(CTRL|VPA)_.*$", "\\1", group_levels),
    donor_id  = sub("^(CTRL|VPA)_", "", group_levels),
    n_cells   = as.integer(table(groups)[group_levels]),
    row.names = group_levels
  )

  list(counts = as.matrix(pb), sample_info = sample_info)
}

pseudobulk_list <- lapply(
  setNames(cell_types_for_de, cell_types_for_de),
  function(ct) make_pseudobulk(obj, ct)
)

# Show sample sizes
pb_summary <- do.call(rbind, lapply(names(pseudobulk_list), function(ct) {
  info <- pseudobulk_list[[ct]]$sample_info[, c("sample", "condition", "n_cells")]
  info$cell_type <- ct
  info
}))
knitr::kable(
  pb_summary[, c("cell_type", "sample", "condition", "n_cells")],
  row.names = FALSE,
  caption = "Pseudobulk sample sizes"
)
```

### limma-voom Differential Expression

We use **limma-voom** with a paired design formula `~donor_id + condition` to
test for VPA effects while blocking on donor variability. The **empirical Bayes**
shrinkage in limma is especially valuable with small sample sizes, as it borrows
information across genes to stabilize variance estimates.

**Important caveat**: With only 2 donors per condition, statistical power is
limited. These results identify candidate genes for follow-up but should not be
treated as definitive.

```{r}
#| label: limma-voom-de
#| message: false
#| warning: false
run_limma_voom <- function(pb, cell_type_name) {
  counts <- pb$counts
  info   <- pb$sample_info

  # Create DGEList
  y <- DGEList(counts = counts, samples = info)

  # Filter lowly expressed genes
  keep <- filterByExpr(y, group = info$condition)
  y <- y[keep, , keep.lib.sizes = FALSE]

  # Normalize
  y <- calcNormFactors(y)

  # Design matrix (paired)
  design <- model.matrix(~donor_id + condition, data = info)

  # voom transformation
  v <- voom(y, design, plot = FALSE)

  # Fit
  fit <- lmFit(v, design)
  fit <- eBayes(fit)

  # Extract results for VPA effect
  # The coefficient name depends on the factor levels
  coef_name <- grep("condition", colnames(coef(fit)), value = TRUE)

  tt <- topTable(fit, coef = coef_name, number = Inf, sort.by = "none")
  tt$gene <- rownames(tt)
  tt$cell_type <- cell_type_name
  tt
}

de_results <- lapply(names(pseudobulk_list), function(ct) {
  run_limma_voom(pseudobulk_list[[ct]], ct)
})
names(de_results) <- names(pseudobulk_list)

de_all <- bind_rows(de_results)
```

### DE Summary

```{r}
#| label: de-summary
de_summary <- de_all |>
  group_by(cell_type) |>
  summarize(
    genes_tested = n(),
    up_FDR05   = sum(adj.P.Val < 0.05 & logFC > 0, na.rm = TRUE),
    down_FDR05 = sum(adj.P.Val < 0.05 & logFC < 0, na.rm = TRUE),
    up_FDR10   = sum(adj.P.Val < 0.10 & logFC > 0, na.rm = TRUE),
    down_FDR10 = sum(adj.P.Val < 0.10 & logFC < 0, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(de_summary,
              caption = "Number of DE genes per cell type (VPA vs CTRL)",
              options = list(dom = "t"))
```

### Volcano Plots

Volcano plots show the relationship between effect size (log fold change, x-axis)
and statistical significance (−log10 adjusted p-value, y-axis). Genes in the
upper corners are both large in effect and statistically significant.

```{r}
#| label: volcano-plots
#| fig-width: 16
#| fig-height: 6
#| warning: false
#| fig-alt: "Volcano plots for each cell type showing log2 fold change versus negative log10 adjusted p-value. Significant genes (FDR < 0.05) are colored red and labeled."
base_size <- 14
volcano_plots <- lapply(names(de_results), function(ct) {
  df <- de_results[[ct]]
  df$sig <- ifelse(df$adj.P.Val < 0.05, "FDR < 0.05", "NS")

  # Top genes to label
  top_genes <- df |>
    dplyr::filter(adj.P.Val < 0.05) |>
    arrange(adj.P.Val) |>
    head(15)

  ggplot(df, aes(logFC, -log10(adj.P.Val), color = sig)) +
    geom_point(size = 0.8, alpha = 0.5) +
    scale_color_manual(values = c("FDR < 0.05" = "red", "NS" = "grey60")) +
    geom_text_repel(
      data = top_genes,
      aes(label = gene),
      size = 0.9 * base_size / ggplot2::.pt, max.overlaps = 20, color = "black"
    ) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
    labs(
      title = ct,
      x = "log2 Fold Change (VPA vs CTRL)",
      y = "-log10(adj. P-value)",
      color = NULL
    ) +
    theme(legend.position = "bottom")
})

wrap_plots(volcano_plots, nrow = 1)
```

### Top DE Genes Table

```{r}
#| label: de-table
#| warning: false
# DT warns "data too big for client-side DataTables" but server-side
# is not available in static HTML; the table renders fine client-side
top_de <- de_all |>
  dplyr::filter(adj.P.Val < 0.10) |>
  arrange(adj.P.Val) |>
  mutate(
    gene = gene_link(gene),
    logFC = round(logFC, 3),
    AveExpr = round(AveExpr, 2),
    t = round(t, 2),
    P.Value = signif(P.Value, 3),
    adj.P.Val = signif(adj.P.Val, 3),
    B = round(B, 2)
  ) |>
  dplyr::select(cell_type, gene, logFC, AveExpr, t, P.Value, adj.P.Val, B)

DT::datatable(
  top_de,
  caption = "DE genes (FDR < 0.10) across cell types — click gene name for GeneCards",
  escape = FALSE,
  options = list(pageLength = 25, order = list(list(6, "asc")))
)
```

### Heatmap of Top DE Genes

```{r}
#| label: de-heatmap
#| fig-width: 12
#| fig-height: 10
#| fig-alt: "Heatmap of top differentially expressed genes across pseudobulk samples, z-scored by log-CPM. Columns are annotated by condition and cell type; rows are genes. Hierarchical clustering groups samples and genes by expression similarity."
# Get top DE genes across all cell types
top_genes_for_heatmap <- de_all |>
  dplyr::filter(adj.P.Val < 0.10) |>
  group_by(cell_type) |>
  arrange(adj.P.Val) |>
  slice_head(n = 25) |>
  ungroup() |>
  pull(gene) |>
  unique()

if (length(top_genes_for_heatmap) > 2) {
  # Build a matrix of logFC values across pseudobulk samples
  # Use the averaged log-CPM from each pseudobulk
  heatmap_data <- list()
  for (ct in names(pseudobulk_list)) {
    pb <- pseudobulk_list[[ct]]
    y <- DGEList(counts = pb$counts, samples = pb$sample_info)
    y <- calcNormFactors(y)
    lcpm <- cpm(y, log = TRUE)
    genes_avail <- intersect(top_genes_for_heatmap, rownames(lcpm))
    if (length(genes_avail) > 0) {
      mat <- lcpm[genes_avail, , drop = FALSE]
      colnames(mat) <- paste0(ct, ":", colnames(mat))
      heatmap_data[[ct]] <- mat
    }
  }

  # Combine and scale
  all_genes <- Reduce(intersect, lapply(heatmap_data, rownames))
  if (length(all_genes) > 2) {
    hm_mat <- do.call(cbind, lapply(heatmap_data, function(m) m[all_genes, ]))
    hm_scaled <- t(scale(t(hm_mat)))

    # Annotation
    ann_col <- data.frame(
      condition = sub(".*:(CTRL|VPA)_.*", "\\1", colnames(hm_mat)),
      cell_type = sub(":.*", "", colnames(hm_mat)),
      row.names = colnames(hm_mat)
    )

    pheatmap(
      hm_scaled,
      annotation_col = ann_col,
      show_rownames = (nrow(hm_scaled) <= 50),
      cluster_cols = TRUE,
      cluster_rows = TRUE,
      main = "Top DE genes across pseudobulk samples (z-scored log-CPM)",
      fontsize_row = 7
    )
  } else {
    cat("Too few shared DE genes across cell types for a meaningful heatmap.\n")
  }
} else {
  cat("Fewer than 3 DE genes at FDR < 0.10; skipping heatmap.\n")
}
```


## Gene Set Enrichment Analysis

Gene Set Enrichment Analysis (GSEA) asks whether genes belonging to known
biological pathways are systematically shifted up or down in expression. Rather
than testing one gene at a time, GSEA tests **sets of genes** that share a
function, making it more sensitive to coordinated but modest changes.

We rank genes by their limma t-statistic (which captures both effect size and
certainty) and test against **Hallmark** and **GO Biological Process** gene sets
from MSigDB.

```{r}
#| label: gsea-gene-sets
#| message: false
# Build TERM2GENE mappings for clusterProfiler
hallmark_t2g <- msigdbr(species = "Homo sapiens", collection = "H") |>
  dplyr::select(gs_name, gene_symbol)

gobp_t2g <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP") |>
  dplyr::select(gs_name, gene_symbol)
```

```{r}
#| label: gsea-run
#| message: false
#| warning: false

# Free Seurat object memory before GSEA — we only need de_results from here
invisible(gc())

# Run Hallmark and GO:BP GSEA separately per cell type to avoid heap
# corruption from very large combined TERM2GENE tables in fgsea
run_gsea_for_ct <- function(ct, t2g) {
  df <- de_results[[ct]]
  gene_list <- setNames(df$t, df$gene)
  gene_list <- sort(gene_list, decreasing = TRUE)
  res <- clusterProfiler::GSEA(
    geneList = gene_list,
    TERM2GENE = t2g,
    minGSSize = 15,
    maxGSSize = 500,
    pvalueCutoff = 1.0,
    verbose = FALSE
  )
  res@result$cell_type <- ct
  res
}

# Hallmark GSEA
gsea_hallmark <- lapply(names(de_results), function(ct) {
  run_gsea_for_ct(ct, hallmark_t2g)
})
names(gsea_hallmark) <- names(de_results)
invisible(gc())

# GO:BP GSEA
gsea_gobp <- lapply(names(de_results), function(ct) {
  run_gsea_for_ct(ct, gobp_t2g)
})
names(gsea_gobp) <- names(de_results)
invisible(gc())

# Merge Hallmark and GO:BP results per cell type into one gseaResult object
gsea_results <- lapply(names(de_results), function(ct) {
  combined <- bind_rows(gsea_hallmark[[ct]]@result, gsea_gobp[[ct]]@result)
  # Return the hallmark object with merged results
  merged <- gsea_hallmark[[ct]]
  merged@result <- combined
  merged
})
names(gsea_results) <- names(de_results)

# Combine results across cell types for tables and custom plots
gsea_all <- bind_rows(lapply(gsea_results, function(r) r@result))
```

### Top Enriched Pathways

```{r}
#| label: gsea-dotplot
#| fig-width: 10
#| fig-height: 10
#| warning: false
#| fig-alt: "Dotplots showing top enriched pathways per cell type from GSEA, with dot size indicating gene set size and color indicating adjusted p-value."
# enrichplot::dotplot for each cell type with significant results
has_sig <- sapply(gsea_results, function(r) sum(r@result$p.adjust < 0.05) > 0)

if (any(has_sig)) {
  for (ct in names(gsea_results)[has_sig]) {
    p <- enrichplot::dotplot(gsea_results[[ct]], showCategory = 15,
                             title = paste("GSEA:", ct))
    print(p)
  }
} else {
  cat("No pathways reached significance at padj < 0.05.\n")
  # Show dotplots at relaxed threshold for exploration
  for (ct in names(gsea_results)) {
    n_nom <- sum(gsea_results[[ct]]@result$pvalue < 0.05)
    if (n_nom > 0) {
      # Temporarily relax cutoff so dotplot has something to show
      res_relaxed <- gsea_results[[ct]]
      res_relaxed@result$p.adjust <- res_relaxed@result$pvalue
      p <- enrichplot::dotplot(res_relaxed, showCategory = 10,
                               title = paste("GSEA (nominal p):", ct))
      print(p)
    }
  }
}
```

### GSEA Results Table

```{r}
#| label: gsea-table
gsea_table <- gsea_all |>
  dplyr::filter(pvalue < 0.05) |>
  arrange(pvalue) |>
  mutate(
    NES = round(NES, 2),
    pvalue = signif(pvalue, 3),
    p.adjust = signif(p.adjust, 3),
    setSize = as.integer(setSize)
  ) |>
  dplyr::select(cell_type, Description, NES, pvalue, p.adjust, setSize) |>
  head(100)

DT::datatable(
  gsea_table,
  caption = "GSEA results (nominal p < 0.05)",
  options = list(pageLength = 25, order = list(list(3, "asc")))
)
```


## Known Biology Check

We examine genes known to be responsive to VPA or involved in neural development
to see whether their expression patterns are consistent with expectations.

**Selected genes**:

- **HDAC targets / chromatin remodeling**: *HDAC1*, *HDAC2*, *HDAC3*, *EP300*, *CREBBP*
- **Neural progenitor markers**: *PAX6*, *SOX2*, *NES*, *HES1*, *NOTCH1*
- **Neuronal differentiation**: *DCX*, *TBR1*, *NEUROD1*, *NEUROD2*, *MAP2*
- **Cortical layer markers**: *SATB2*, *BCL11B* (CTIP2), *TLE4*
- **Apoptosis / cell stress**: *TP53*, *BCL2*, *BAX*

```{r}
#| label: known-genes-data
known_genes <- c(
  "HDAC1", "HDAC2", "HDAC3", "EP300", "CREBBP",
  "PAX6", "SOX2", "NES", "HES1", "NOTCH1",
  "DCX", "TBR1", "NEUROD1", "NEUROD2", "MAP2",
  "SATB2", "BCL11B", "TLE4",
  "TP53", "BCL2", "BAX"
)

# Filter to genes present in data
known_genes_present <- known_genes[known_genes %in% rownames(obj)]
knitr::kable(
  data.frame(
    Status = c("Present in dataset", "Missing"),
    Count = c(length(known_genes_present), length(known_genes) - length(known_genes_present)),
    Genes = c(
      paste(known_genes_present, collapse = ", "),
      paste(setdiff(known_genes, known_genes_present), collapse = ", ")
    )
  ),
  row.names = FALSE,
  caption = "Known gene availability"
)
```

### Violin Plots

```{r}
#| label: known-genes-violin
#| fig-width: 14
#| fig-height: 8
#| warning: false
#| fig-alt: "Violin plots showing expression of known VPA-response and neural development genes across major cell types, split by CTRL and VPA conditions. Each panel shows one gene in one cell type."
# Extract expression for known genes using Seurat's FetchData
if (length(known_genes_present) > 0) {
  expr_df <- FetchData(obj, vars = c(known_genes_present, "condition", "author_cell_type"))
  colnames(expr_df)[colnames(expr_df) == "author_cell_type"] <- "cell_type"

  expr_long <- expr_df |>
    tidyr::pivot_longer(
      cols = all_of(known_genes_present),
      names_to = "gene",
      values_to = "expression"
    )

  # Only plot for the major cell types
  expr_long_major <- expr_long |>
    dplyr::filter(cell_type %in% cell_types_for_de)

  # Split genes into groups of 4 to keep panels readable (4 genes × 5 cell types = 20 panels)
  genes_per_plot <- 4
  gene_groups <- split(
    unique(expr_long_major$gene),
    ceiling(seq_along(unique(expr_long_major$gene)) / genes_per_plot)
  )

  for (i in seq_along(gene_groups)) {
    p <- expr_long_major |>
      dplyr::filter(gene %in% gene_groups[[i]]) |>
      ggplot(aes(condition, expression, fill = condition)) +
      geom_violin(scale = "width", alpha = 0.7, linewidth = 0.3) +
      scale_fill_manual(values = condition_colors) +
      facet_grid(gene ~ cell_type, scales = "free_y", switch = "y") +
      theme(
        strip.placement = "outside",
        strip.text.y.left = element_text(angle = 0, size = rel(0.7)),
        strip.text.x = element_text(size = rel(0.75)),
        axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7))
      ) +
      labs(
        title = paste0(
          "VPA-response and neural genes (group ", i,
          " of ", length(gene_groups), ")"
        ),
        x = NULL, y = "log1p expression", fill = "Condition"
      )
    print(p)
  }
}
```

### Feature Plots (UMAP) for Top DE Genes

```{r}
#| label: feature-plots
#| fig-width: 16
#| fig-height: 12
#| warning: false
#| fig-alt: "UMAP feature plots showing expression of top differentially expressed genes. Each panel shows one gene, with cells colored by expression level from grey (low) to dark magenta (high)."
# Pick top DE genes across cell types
top_de_genes <- de_all |>
  dplyr::filter(adj.P.Val < 0.10) |>
  arrange(adj.P.Val) |>
  pull(gene) |>
  unique() |>
  head(8)

# If too few DE genes, use top by nominal p
if (length(top_de_genes) < 4) {
  top_de_genes <- de_all |>
    arrange(P.Value) |>
    pull(gene) |>
    unique() |>
    head(8)
}

if (length(top_de_genes) > 0) {
  # Filter to genes present in the Seurat object
  top_de_genes <- top_de_genes[top_de_genes %in% rownames(obj)]
  FeaturePlot(obj, features = top_de_genes, reduction = "author_umap",
              cols = c("lightgrey", "darkmagenta"), ncol = 4, pt.size = 0.2)
}
```


## Draft Results

```{r}
#| label: results-data
#| echo: false
n_cells <- ncol(obj)
n_ctrl <- sum(obj$condition == "CTRL")
n_vpa <- sum(obj$condition == "VPA")
n_donors <- length(unique(obj$donor_id))
n_cell_types <- length(unique(obj$author_cell_type))

# Composition changes
comp_summary <- comp_cond |>
  dplyr::select(condition, cell_type, proportion) |>
  tidyr::pivot_wider(names_from = condition, values_from = proportion) |>
  mutate(change_pct = round((VPA - CTRL) * 100, 1))

# DE counts
n_de_005 <- sum(de_all$adj.P.Val < 0.05, na.rm = TRUE)
n_de_010 <- sum(de_all$adj.P.Val < 0.10, na.rm = TRUE)

# Top pathway
top_pw <- gsea_all |> arrange(pvalue) |> head(1)
```

### Results Summary

We analyzed `r format(n_cells, big.mark = ",")` cells from
`r n_donors` donors across `r n_cell_types` cell types, comparing CTRL
(`r format(n_ctrl, big.mark = ",")` cells) to VPA-treated
(`r format(n_vpa, big.mark = ",")` cells) human dorsal forebrain organoids.

**Cell type composition**: VPA treatment was associated with notable shifts in
cell type proportions:

```{r}
#| label: results-comp-table
#| echo: false
DT::datatable(
  comp_summary |>
    mutate(
      CTRL = paste0(round(CTRL * 100, 1), "%"),
      VPA = paste0(round(VPA * 100, 1), "%"),
      `Change (pp)` = change_pct
    ) |>
    dplyr::select(cell_type, CTRL, VPA, `Change (pp)`),
  caption = "Cell type proportion changes (percentage points)",
  options = list(dom = "t"),
  rownames = FALSE
)
```

**Differential expression**: Across the three major cell types tested,
we identified `r n_de_005` genes at FDR < 0.05 and `r n_de_010` genes at
FDR < 0.10.

**Pathway enrichment**: The top enriched pathway was
*`r gsub("_", " ", top_pw$Description[1])`* in `r top_pw$cell_type[1]`
(NES = `r round(top_pw$NES[1], 2)`, nominal p = `r signif(top_pw$pvalue[1], 3)`).

**Limitations**: With only `r n_donors` donors per condition, these results
are exploratory. The use of back-transformed log1p data rather than raw counts
introduces approximation, though pseudobulk aggregation mitigates this.


## Draft Methods

### Data Source

Single-cell RNA-seq data were obtained from CELLxGENE
(dataset ID: `3a8aec06-3309-4d37-b75d-c59f5f6d55a6`), originating from
Yentür et al. (2025, bioRxiv). The dataset comprises human dorsal forebrain
organoids from `r n_donors` donors under control and VPA-treated conditions
(`r format(n_cells, big.mark = ",")` cells total).

### Preprocessing

Data were provided as log1p-normalized expression values and converted to a
Seurat v5 object. Approximate raw counts were obtained by back-transforming
with `expm1()` and rounding. UMAP coordinates and cell type annotations from
the original authors were used without modification. De novo clustering was
performed using Seurat's standard pipeline: `FindVariableFeatures` (2,000 HVGs
via variance-stabilizing transformation), `ScaleData`, `RunPCA` (30 PCs),
`FindNeighbors` and `FindClusters` (Louvain algorithm on an SNN graph, first
20 PCs), and `RunUMAP` (Hao et al. 2024).

### Differential Expression

Pseudobulk expression profiles were generated by summing back-transformed
(expm1) approximate counts per gene across cells within each donor × condition
× cell type group. Differential expression was performed using the limma-voom
pipeline (Ritchie et al. 2015) with a paired design formula (~donor_id +
condition). Genes were filtered using `filterByExpr()`, library sizes
normalized with TMM (`calcNormFactors()`), and precision weights estimated
via `voom()`. Empirical Bayes moderation was applied with `eBayes()`.
P-values were adjusted for multiple testing using the Benjamini-Hochberg method.

### Pathway Analysis

Gene set enrichment analysis was performed using clusterProfiler (Wu et al. 2021),
which uses fgsea (Korotkevich et al. 2021) internally, with genes ranked by limma
t-statistic. MSigDB Hallmark and GO Biological Process gene sets were tested
(gene set sizes 15–500). Visualization used enrichplot.

### Software

```{r}
#| label: methods-software
#| echo: false
#| results: asis
pkgs <- c("R", "Seurat", "SingleCellExperiment", "edgeR", "limma",
           "clusterProfiler", "enrichplot", "msigdbr", "ggplot2", "plotly")
versions <- sapply(pkgs, function(p) {
  if (p == "R") {
    paste(R.version$major, R.version$minor, sep = ".")
  } else {
    as.character(packageVersion(p))
  }
})
cat(paste(
  paste0("**", pkgs, "** v", versions),
  collapse = ", "
), "\n")
```


## References

- Hao Y et al. (2024) "Dictionary learning for integrative, multimodal and scalable single-cell analysis." *Nat Biotechnol* 42:293–304. <https://doi.org/10.1038/s41587-023-01767-y>
- Ritchie ME et al. (2015) "limma powers differential expression analyses for RNA-sequencing and microarray studies." *Nucleic Acids Res* 43(7):e47. <https://doi.org/10.1093/nar/gkv007>
- Law CW et al. (2014) "voom: precision weights unlock linear model analysis tools for RNA-seq read counts." *Genome Biol* 15:R29. <https://doi.org/10.1186/gb-2014-15-2-r29>
- Robinson MD, McCarthy DJ, Smyth GK (2010) "edgeR: a Bioconductor package for differential expression analysis of digital gene expression data." *Bioinformatics* 26(1):139–140. <https://doi.org/10.1093/bioinformatics/btp616>
- Wu T et al. (2021) "clusterProfiler 4.0: A universal enrichment tool for interpreting omics data." *Innovation* 2(3):100141. <https://doi.org/10.1016/j.xinn.2021.100141>
- Korotkevich G et al. (2021) "Fast gene set enrichment analysis." *bioRxiv*. <https://doi.org/10.1101/060012>
- Liberzon A et al. (2015) "The Molecular Signatures Database Hallmark Gene Set Collection." *Cell Syst* 1(6):417–425. <https://doi.org/10.1016/j.cels.2015.12.004>
- Amezquita RA et al. (2020) "Orchestrating single-cell analysis with Bioconductor." *Nat Methods* 17:137–145. <https://doi.org/10.1038/s41592-019-0654-x>
- Squair JW et al. (2021) "Confronting false discoveries in single-cell differential expression." *Nat Commun* 12:5692. <https://doi.org/10.1038/s41467-021-25960-2>
- McInnes L, Healy J, Melville J (2018) "UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction." *arXiv:1802.03426*. <https://doi.org/10.48550/arXiv.1802.03426>


::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info
#| echo: false
#| comment: ""
sessioninfo::session_info(info = "all")
```
:::
